---
title: "Spatial Point Pattern Analysis"
author: "Federico Jose Rodriguez"
date: "Sep 2 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

# Getting Started

## Data Sources

The exercise uses the data sources to be used in the upcoming Take Home Exercise:

-   Thailand Road accident data from 2019-2022 from [Kaggle](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022)

-   Thailand Roads OpenStreetMap from [HDX](https://data.humdata.org/dataset/hotosm_tha_roads)

-   Thailand - Subnational Administrative Boudaries shapefile from [HDX](https://data.humdata.org/dataset/cod-ab-tha?)

## Installing and launching R packages

This exercise will make use of four R packages: **sf**, **spatstat,** **tidyverse, maptools** and **tmap**.

The code chunk below imports the already retired. We can still download it from Posit Public Package Manager snapshots by using the code below.

```{r}
#| eval: false
install.packages("maptools",
                 repos = "https://packagemanager.posit.com/cran/2023-10-13")
```

The code chunk below uses `p_load()` of **pacman** package to check if the packages are installed in the computer. It installs them first if they are not. It then loads them into R.

```{r}
pacman::p_load(sf, tidyverse, tmap, spatstat, maptools)
```

# Data Importation and Data Wrangling

## Importing the Aspatial Data and converting to sf

The Thailand road accident data is in csv format but contains a field for longitude and another for latitude.

```{r}
rdacc_sf <- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")  %>%
  filter(!is.na(longitude) & longitude != "",
         !is.na(latitude) & latitude != "") %>%
  st_as_sf(coords = c(
    "longitude", "latitude"),
    crs=4326) %>%
  st_transform(crs = 32647)
```

The data appears to have been loaded properly. With the code chunk below, we confirm the data is loaded with a little loss of data given we have filtered out records with invalid coordinates.

```{r}
rdacc_sf
```
