---
title: "Geographically Weighted Regression Model (sf methods)"
author: "Federico Jose Rodriguez"
date: "Oct 14 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

In this in-class exercise, we learn to use GWR or geographically weighted regression. We use sf methods unlike the ones used in the corresponding hands-oon exercise

# Getting Started

## Data Sources

To datasets will be used for this exercise:

-   2014 Master Plan subzone boundary in shapefile format

-   2015 condo resale prices in csv format

## Installing and launching R packages

This exercise will make use of the following R packages:

-   **olsrr -** for building OLS (ordinary least squares) regression models and performing diagnostic tests

-   **GWmodel -** for calibrating geographically weighted family of models

-   **tmap -** for plotting cartographic quality maps

-   **corrplot** - for multivariate data visualization and analysis

-   **sf** - spatial data handling

-   **tidyverse** - attribute data handling

The code chunk below uses `p_load()` of **pacman** package to check if the packages are installed in the computer. It installs them first if they are not. It then loads them into R.

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary, ggstatsplot, sfdep)
```

# Data Import and Preparation

## Geospatial data loading and preparation

The code chunk below uses `st_read()` of the **sf** package to load the geospatial data and apply the right EPSG code to convert it to svy21.

```{r}
mpsz_svy21 = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL") %>%
  st_transform(3414) %>%
  st_make_valid()
```

## Aspatial data loading

The code chunk below uses `read_csv()` of **readr** to import the 2015 condo resale prices from the csv file.

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
```

## Converting aspatial dataframe into an sf object

To convert the `condo_resale` object into a spatial object, we can use the following code chunk that utilizes `st_as_sf()` from sf package. The final line of the code chunk converts the data frame from wgs84 to svy21 using the indicated crs values.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

# Exploratory Data Analysis

For the in-class exercise, we focus on analysing if there are any highly correlated variables in the dataset.

## Checking for correlation

In the hands-on exercise, we used `corrplot()` to generate the correlation plot.

An alternative approach is to use `ggcorrmat()` of **ggstatsplot** which requires a simpler line of code.

```{r fig.width=12, fig.height=10}
ggcorrmat(condo_resale[, 5:23])
```

# Hedonic Pricing Model

We run the global model using `lm()`. The code chunk below includes all variables.

```{r}
condo.mlr <- lm(formula = SELLING_PRICE ~ AREA_SQM + AGE    + 
                  PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                  PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET +
                  PROX_KINDERGARTEN + PROX_MRT  + PROX_PARK +
                  PROX_PRIMARY_SCH + PROX_TOP_PRIMARY_SCH +
                  PROX_SHOPPING_MALL + PROX_SUPERMARKET + PROX_BUS_STOP +
                  NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD + LEASEHOLD_99YR, 
                data=condo_resale.sf)
summary(condo.mlr)
```

The following code gives the results of the model.

```{r}
summary(condo.mlr)
```

The mlm object also has some useful objects like `coefficients` which contains the variable coefficients.

```{r}
condo.mlr$coefficients
```

## Using olsrr to display results and for VIF

We can use the olsrr package to run diagnostic tests on the model.

We can pass the results to `ols_regress()` to produce a formatted model report, better than the one coming from `summary()`.

```{r}
ols_regress(condo.mlr)
```

We can then use `ols_vif_tol()` to run the test for multicollinearity using the VIF or variance inflation factor.

```{r}
filter(ols_vif_tol(condo.mlr), VIF > 5) # sign of multicollinearity
filter(ols_vif_tol(condo.mlr), VIF <= 5) # no sign of multicollinearity
```

As there is no variable with VIF value above 5, we are ensured that there is no signs of multicollinearity using this criterion.

## Using olsrr for variable selection using stepwise regression

Forward stepwise regression adds variables one at a time from an empty model by picking the one with the highest rank for a criteria (and within a threshold) and that improves the model. (i.e., adj R squared) The criteria is typically the significance level (e.g., choose minimum p-value below 0.05)

`ols_step_forward_p()` performs forward stepwise regression using the p-value. Other criteria include AIC, BIC, r-squared. A 0.05 max p-value is defined using the `p_val` argument. The `details` argument instructs whether the results are printed out while each step is run.

```{r}
condo_fw_mlr <- ols_step_forward_p(condo.mlr, p_val = 0.05, details = FALSE)
```

We can pass the ols regression object into `plot()` to display the results graphically. The charts show the improvement of four of the model metrics with each variable added.

```{r}
#| fig-width: 12
#| fig-height: 10
plot(condo_fw_mlr)
```

We can plot the residuals using `ols_plot_resid_fit()` and then passing the `model` object inside the mlr object. This tests for the linearity assumption.

```{r}
ols_plot_resid_fit(condo_fw_mlr$model)
```

We can plot the residuals using `ols_plot_resid_hist()` to test for the normality assumption graphically.

```{r}
ols_plot_resid_hist(condo_fw_mlr$model)
```

We can also do it using the traditional stats using the following

```{r}
ols_test_normality(condo_fw_mlr$model)
```

## Testing spatial autocorrelation

We export the residuals of the hedonic pricing model as a dataframe

```{r}
mlr_output <- as.data.frame(condo_fw_mlr$model$residuals) %>%
  rename('FW_MLR_RES' = 'condo_fw_mlr$model$residuals')
```

We then join the new dataframe to the sf object.

```{r}
condo_resale.sf <- cbind(condo_resale.sf,
                         mlr_output$FW_MLR_RES) %>%
  rename('MLR_RES' = 'mlr_output.FW_MLR_RES')
```

With this, we can produce an interactive map of the residuals using the sf object.

```{r}
tmap_mode("view")
tmap_options(check.and.fix = TRUE) # can be added to the layer with a problem

tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.4) +
tm_shape(condo_resale.sf) +  
  tm_dots(col = "MLR_RES",
          alpha = 0.6,
          style="quantile", title = "Residual") +
  tm_view(set.zoom.limits = c(11,14))

tmap_mode("plot")
```

There appears to be clusters with high residuals– there appears to be signs of spatial autocorrelation.

To prove our suspicions, we conduct Moran's I using the sfdep package (without needing to convert and use spdep as in the hands-on exercise)

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(nb = st_knn(geometry, k = 6,
                     longlat = FALSE),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

We can then run the Global Moran's I permutation test directly on the object

```{r}
global_moran_perm(condo_resale.sf$MLR_RES,
                  condo_resale.sf$nb,
                  condo_resale.sf$wt,
                  alternative = "two.sided",
                  nsim = 99)
```

Since the test p-value is less than 0.05, the result is significant, and as the test statistic is positive, then there are signs of clusters.

As there is spatial autocorrelation, we can build the GWR model.

# Building GWR Model

## Using fixed bandwidth method

The following code chunk derives the optimal fixed bandwidth for the model from the regression method (have to type in full as it cannot take in the ols regression object.

```{r}
bw_fixed <- bw.gwr(formula = SELLING_PRICE ~ AREA_SQM + AGE    + 
                  PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                  PROX_URA_GROWTH_AREA + PROX_MRT  + PROX_PARK +
                  PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
                  NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
                  data = condo_resale.sf,
                  approach = "CV",
                  kernel = "gaussian",
                  adaptive = FALSE,
                  longlat = FALSE)
```

The results show that \~971.3405 is the recommended bandwidth.

The code below calibrates the gwr model using fixed bandwidth

```{r}
gwr_fixed <- gwr.basic(formula = SELLING_PRICE ~ AREA_SQM + AGE    + 
                  PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                  PROX_URA_GROWTH_AREA + PROX_MRT  + PROX_PARK +
                  PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
                  NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
                  data = condo_resale.sf,
                  bw = bw_fixed,
                  kernel = "gaussian",
                  longlat = FALSE)
```

The output is saved in a list of class gwrm

```{r}
gwr_fixed
```

The resulting object has a field named SDF which contains the parameter estimates. To visualize elements of it, we transfer this information into the sf object. (only considering a few columns)

```{r}
gwr_fixed_output <- as.data.frame(gwr_fixed$SDF) %>%
  select(-c(2:15))

gwr_sf_fixed <- cbind(condo_resale.sf, gwr_fixed_output)
```

We can use `glimpse()` to check the contents of the new object

```{r}
glimpse(gwr_sf_fixed)
```

We can then plot using tmap

```{r}
tmap_mode("view")

tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(gwr_sf_fixed) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_mode("plot")
```
