---
title: "Geographically Weighted Modeling of Financial Inclusion in Tanzania"
author: "Federico Jose Rodriguez"
date: "Oct 21 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

WIP.

# A. Getting Started

## A.1 Background

WIP.

Define financial inclusion

Describe state and statistics in Africa and Tanzania

## A.2 Objectives

WIP.

## A.3 Data Sources

The following data sources are used for this analysis:

-   XXX from [Finscope Tanzania](https://www.fsdt.or.tz/finscope/)

-   District-level boundaries in Tanzania as a shapefile from [geoBoundaries.org](https://www.geoboundaries.org/) portal

## A.4 Importing and Launching R Packages

For this study, the following R packages will be used. A description of the packages and the code, using `p_load()` of the **pacman** package, to import them is given below.

::: panel-tabset
#### Package Description

The loaded packages include:

-   **sf -** package for importing, managing and processing vector-based geospatial data

-   **tidyverse -** collection of packages for performing data importation, wrangling and visualization

-   **tmap -** package with functions for plotting cartographic quality maps

-   **sfdep** - for handling spatial data

-   **coorplot, ggpubr, heatmaply, factoextra** - packages for multivariate data visualization and analysis

-   **cluster, ClustGeo, NbClust** - packages for performing cluster analysis

#### Import Code

```{r}
pacman::p_load(sf, tmap, spdep, sfdep, tidyverse,
               ggpubr, heatmaply, factoextra,
               NbClust, cluster, ClustGeo)
```
:::

As we will be performing simulations in the analysis later, it is good practice to define a random seed to be used so that results are consistent for viewers of this report, and the results can be reproduced.

```{r}
set.seed(1234)
```

# B. Data Loading and Preparation

## B.1. Tanzania District boundaries

TBC.

```{r}
tz_dist <- st_read(dsn="data/geospatial", 
                   layer="geoBoundaries-TZA-ADM3")
```

xxx

```{r}
qtm(tz_dist)
```
