---
title: "Second Order Spatial Point Pattern Analysis Methods"
author: "Federico Jose Rodriguez"
date: "Sep 1 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

In this hands-on exercise, we continue learning about Spatial Point pattern analysis with the help of **spatstat** package. We continue looking at the dataset using the location of childcare centres in Singapore.

This exercise is based on Chapter 5 of Dr Kam's online book which can be accessed [here](https://r4gdsa.netlify.app/ "R for Geospatial Data Science and Analytics by Dr Kam").

# Getting Started

## Data Sources

Data for this exercise are from public sources and are the same as the previous hands-on exercise. This includes:

-   Location and attribute information of childcare centres in Singapore from data.gov.sg

-   Master Plan 2014 Subzone Boundary from [data.gov.sg](https://beta.data.gov.sg/datasets/d_d14da225fccf921049ab64238ff473d9/view)

-   National boundary of Singapore provided in SLA and ESRI shapefile format

## Installing and launching R packages

This exercise will make use of five R packages: **sf**, **tidyverse, spatstat, raster** and **tmap**.

-   **sf -** for importing, managing and processing vector-based geospatial data

-   **tidyverse -** collection of packages for performing data importation, wrangling and visualization

-   **tmap -** for plotting cartographic quality maps

-   **spatstat** - offers a wide range of functions for point pattern analysis (PPA)

-   **raster** - for reading, writing, manipulating and analysing models of gridded spatial data

The code chunk below uses `p_load()` of **pacman** package to check if the packages are installed in the computer. It installs them first if they are not. It then loads them into R.

```{r}
pacman::p_load(sf, raster, spatstat, tmap, tidyverse)
```

# Spatial Data Wrangling

## Importing the spatial data

The code chunk below uses `st_read()` from the **sf** package to import the geospatial datasets.

```{r}
childcare_sf <- st_read("data/geospatial/child-care-services-geojson.geojson") %>% st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data/geospatial", layer="CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

## Checking and converting projection systems

We can use `st_crs()` to check what coordinate systems are used in each of the three sf dataframes.

`childcare_sf` is in SVY21 using EPSG code 3414 after our load and transform operation above. However, `sg_sf` and `mpsz_sf` reveals that they are not using the correct EPSG code.

```{r}
#| echo: false
st_crs(childcare_sf)
```

```{r}
st_crs(sg_sf)
```

```{r}
st_crs(mpsz_sf)
```

We use the code chunk below to update the EPSG code for the last two objects

```{r}
sg_sf = st_set_crs(sg_sf, 3414)
mpsz_sf = st_set_crs(mpsz_sf, 3414)
```

## Mapping the geospatial datasets

It is always good practice to plot or display imported data to check that they have been loaded and transformed properly. The code chunk below creates a map using tmap with all three objects where they appear to be mapped properly.

```{r}
tm_shape(sg_sf)+
    tm_fill("lightblue") +
    tm_borders(lwd = 0.1,  alpha = 1)+
    tm_shape(mpsz_sf) +
    tm_fill("grey", alpha = 0.5) +
    tm_borders(lwd = 0.1,  alpha = 1) +
    tm_shape(childcare_sf) +
    tm_dots(col = "darkgreen")
```

# Geospatial Data Wrangling

## Converting sf dataframe sp spatial class

The code chunk below converts the three dataframes using `as_Spatial()` from the **sf** package.

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

We can check the contents of the new dataframes by calling them. This confirms that they are in the spatial\* class format.

```{r}
childcare
```

## Converting the spatial class to generic sp format

**spatstat** requires that the data is in ***ppp*** object form. There is no direct way to do this from spatial\* class. We need to convert spatial\* class to a spatial object first.

The code chunk below transform two of the spatial\* objects into generic sp objects.

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

Calling `childcare_sp` and `sg_sp` lets us check their properties.

```{r}
childcare_sp
```

```{r}
sg_sp
```

## Converting generic sp format into spatstat's ppp format

We will then use `as.ppp()` from **spatstat** package to convert the (spatial) data into spatstat's ppp object format.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
childcare_ppp
```

We see the difference of this format when we use `plot()` (from R Graphics) to produce a quick map of the data.

```{r}
plot(childcare_ppp)
```

We can see summary information on the new ppp object using the code chunk below.

```{r}
summary(childcare_ppp)
```

## Handling duplicated points

We can check duplication in a ppp object using the code chunk below.

```{r}
any(duplicated(childcare_ppp))
```

To count the number of coincident points, we can use the `multiplicity()` function.

```{r}
multiplicity(childcare_ppp)
```

We can wrap this in `sum()` to count the number of locations with more than one event.

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

The outputs show that there are no duplication in `childcare_ppp`

## Creating owin object

When analysing spatial point patterns, it is best to confine the analysis within a geographical area. In **spatstat**, the object that represents the bounded region is called an **owin**.

The code chunk below creates an **owin** based on the `sg` SpatialPolygon object.

```{r}
sg_owin <- as.owin(sg_sf)
```

The owin object can be displayed graphically using `plot()` and summarized using `summary()`

```{r}
plot(sg_owin)
```

```{r}
summary(sg_owin)
```

## Combining point events object and owin object

In this last step, we extract childcare events (locations) that are within Singapore, as depicted by the owin, using the code chunk below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

The output is a combination of the point feature and the polygon feature into a single ppp object:

```{r}
summary(childcareSG_ppp)
```

Running `plot()` on this shows both the owin and the preschool locations in a single map.

```{r}
plot(childcareSG_ppp)
```

## Extracting study areas

The code chunk below extracts the target planning areas from `mpsz_sf`

```{r}
pg <- mpsz_sf %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_sf %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_sf %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_sf %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

The code chunks below plot the different target planning areas using `tmap_arrange()`.

```{r}
punggol <- tm_shape(pg) + 
  tm_polygons() + tm_layout(title = "Punggol")

cck <- tm_shape(ck) + 
  tm_polygons() + tm_layout(title = "Chua Chu Kang")

tampines <- tm_shape(tm) + 
  tm_polygons() + tm_layout(title = "Tampines")

jwest <- tm_shape(jw) + 
  tm_polygons() + tm_layout(title = "Jurong West")

tmap_arrange(punggol, tampines, cck, jwest, asp=2, ncol=2, nrow = 2)
```

## Converting target areas into owin objects

The code chunk below converts the four objeects into owin which is a requirement for analysing with **spatstat**

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

## Combining childcare points and the study area

The code chunk below extracts the childcare points/events that are within each of the target areas

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

We use `rescale.ppp()` in the next code chunk to transform the unit of measure from meter to kilometer

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

The code chunk below is used to plot the four target areas and the locations of their childcare centres

```{r}
par(mfrow=c(2,2))
plot(childcare_pg_ppp.km, main="Punggol")
plot(childcare_tm_ppp.km, main="Tampines")
plot(childcare_ck_ppp.km, main="Choa Chu Kang")
plot(childcare_jw_ppp.km, main="Jurong West")
```

# Second-order Spatial Point Patterns Analysis 

## Analysing spatial point process using G function

The G function measures the distribution of distances from an arbitrary event to the nearest event. We will compute the G function using `Gest()` of the **spatstat** package. We will also perform monte carlo simulation using the `envelope()` function of the same package.

### G-Function Estimation for Choa Chu Kang Planning Area

#### Computing the G function estimation

The code chunk below computes the G-function for the planning area using `Gest()`

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

#### Performing Complete Spatial Randomness (CSR) test

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Choa Chu Kang are randomly distributed

-   $H_1$ - The locations of childcare service centres in Choa Chu Kang are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the G-function values using `envelope()`

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

We then plot the results using the code chunk below

```{r}
plot(G_CK.csr)
```

### G Function Estimation for Tampines Planning Area

#### Computing the G function estimation

The code chunk below computes the G-function for the planning area using `Gest()`

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

#### Performing Complete Spatial Randomness (CSR) test

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Tampines are randomly distributed

-   $H_1$ - The locations of childcare service centres in Tampines are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the G-function values using `envelope()`

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

## Analysing Spatial Point Process using F-Function

The F-Function estimates the empty space function F(r) or its hazard rate h(r) from a point pattern in a window of arbitrary shape. In this section, we will compute the F-function using `Fest()` of the spatstat package. We will again use monte carlo simulation using `envelope()`

### F-Function Estimation for Choa Chu Kang Planning Area

#### Computing the F-function estimation

The code chunk below computes the F-function for the planning area using `Fest()`

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

#### Performing Complete Spatial Randomness (CSR) test

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Choa Chu Kang are randomly distributed

-   $H_1$ - The locations of childcare service centres in Choa Chu Kang are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the F-function values using `envelope()`

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

### F-Function Estimation for Tampines Planning Area

#### Computing the F-function estimation

The code chunk below computes the F-function for the planning area using `Fest()`

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

#### Performing Complete Spatial Randomness (CSR) test

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Tampines are randomly distributed

-   $H_1$ - The locations of childcare service centres in Tampines are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the F-function values using `envelope()`

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

## Analysing Spatial Point Process using K-Function

The K-function measures the number of events found up to a given distance of any particular event. In this section, we will use `Kest()` of the **spatstat** package to estimate the K-function. We will again perform monte carlo simulations using `envelope()`

### K-Function Estimation for Choa Chu Kang Planning Area

The code chunk below computes the K-function for the planning area using `Kest()`

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Choa Chu Kang are randomly distributed

-   $H_1$ - The locations of childcare service centres in Choa Chu Kang are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the K-function values using `envelope()`

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### K-Function Estimation for Tampines Planning Area

The code chunk below computes the K-function for the planning area using `Kest()`

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Tampines are randomly distributed

-   $H_1$ - The locations of childcare service centres in Tampines are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the K-function values using `envelope()`

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## Analysing Spatial Point Process using L-Function

In this section, we will use `Lest()` of the **spatstat** package to estimate the L-function. We will again perform monte carlo simulations using `envelope()`

### L-Function Estimation for Choa Chu Kang Planning Area

The code chunk below computes the L-function for the planning area using `Lest()`

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Choa Chu Kang are randomly distributed

-   $H_1$ - The locations of childcare service centres in Choa Chu Kang are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the L-function values using `envelope()`

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

### L-Function Estimation for Tampines Planning Area

The code chunk below computes the L-function for the planning area using `Lest()`

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

To confirm the observed spatial pattern, we conduct hypothesis testing. The test hypotheses for these are:

-   $H_0$ - The locations of childcare service centres in Tampines are randomly distributed

-   $H_1$ - The locations of childcare service centres in Tampines are **not** randomly distributed

For our testing, we will use a p-value smaller than α = 0.001 to reject the null hypothesis.

The code chunk below produces monte carlo simulation for the L-function values using `envelope()`

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```
