---
title: "Geospatial Analysis for Public Good: A Data-driven Perspective on Road Traffic Accidents in the Bangkok Metropolitan Region"
author: "Federico Jose Rodriguez"
date: "Sep 8 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

(Summary here. WIP)

# A. Getting Started

## A.1 Background

Road traffic accidents account for 1.19million deaths and up to 50 million non-fatal injuries according to a [report by the WHO](https://www.who.int/news-room/fact-sheets/detail/road-traffic-injuries) last year.

The same report identifies major risk groups: low- and middle-income countries, (esp in Africa and Europe) the working population, and males. It also identifies some key risk factors which include human error, speeding, driving under the influence of alcohol, distracted driving, unsafe road infrastructure, unsafe vehicles, and law enforcement. Most of the factors identified are behavioral in nature but do not discount that other factors may also contribute to a higher risk of occurrence.

Within Southeast Asia, Thailand has ranked the highest in terms of incidence of road traffic accidents with an average number of of 20,000 deaths a year or 56 a day. The country has also seen an increase in the number of accidents from 2014 to 2021. A large 19% of these accidents occurred in national highways, and the chances of encountering an accident-prone zone was found to be 66%.

## A.2 Objectives

This study aims to take a deeper look into the road accidents in Thailand, focusing on the [Bangkok Metropolitan Region (BMR)](https://en.wikipedia.org/wiki/Bangkok_Metropolitan_Region) which contains the capital Bangkok, and five neighboring provinces. (Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, Samut Sakhon)

As most literature has focused on behavioral and environmental factors, the study will focus on identifying spatiotemporal factors influencing the occurrence of road accidents in BMR. At the minimum, the study deliverables include the following:

-   Visualization of spatiotemporal dynamics of road traffic accidents in BMR

-   Detailed spatial analysis of road traffic accidents in BMR

-   Detailed spatiotemporal analysis of road traffic accidents in BMR

The appropriate technique must be used for these deliverables and all the analysis and visualizations will be carried out using R.

## A.3 Data Sources

The study makes use of the following datasets which are publicly available online.

| Dataset Short Name | Description                                               | Datasource                                                                              |
|------------------|----------------------|--------------------------------|
| THRA               | Thailand road accident data from 2019 to 2022             | [Kaggle](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022) |
| THOSM              | Thailand roads open street map in shapefile format        | [HDX](https://data.humdata.org/dataset/hotosm_tha_roads)                                |
| THSAB              | Thailand - Subnational Administrative Boudaries shapefile | [HDX](https://data.humdata.org/dataset/cod-ab-tha?)                                     |
|                    |                                                           |                                                                                         |

## A.4 Importing and Launching R Packages

For this study, five R packages will be used. A description of the packages and the code, using `p_load()` of the **pacman** package, to import them is given below.

::: panel-tabset
#### Package Description

The loaded packages include:

-   **sf -** package for importing, managing and processing vector-based geospatial data

-   **tidyverse -** collection of packages for performing data importation, wrangling and visualization

-   **tmap -** package with functions for plotting cartographic quality maps

-   **sPNetwork** - provides functions for performing SPPA methods like KDE and K-function on a network. The package can also be used to build spatial matrices to conduct traditional spatial analyses with spatial weights based on reticular distances

-   **spatstat** - offers a wide range of functions for point pattern analysis (PPA)

-   **raster** - for reading, writing, manipulating and analysing models of gridded spatial data

#### Import Code

```{r}
pacman::p_load(sf, spNetwork, tmap, tidyverse, spatstat, raster)
```
:::

As we will be performing simulations in the analysis later, it is good practice to define a random seed to be used so that results are consistent for viewers of this report, and the results can be reproduced.

```{r}
set.seed(1234)
```

# B. Data Loading and Preparation

In this section, the appropriate R packages will be used to load the three datasets into R objects, and to inspect and transform them as needed.

## B.1 THSAB (Thailand Subnational Administrative Boundary, Shapefile)

We load the Thailand subnational administrative boundary shapefile into an R dataframe using `st_read()` from the sf package. The source provides the geospatial data in varying levels as indicated by their suffix: country (0), province (1), district (2), and sub-district. (3) For focusing on the BMR, which covers Bangkok and neighboring provinces, province is the most likely level of detail we will need so we will use the code chunk below to load the appropriate layer first.

```{r}
thsab_prov <- st_read(dsn="data/geospatial", 
                   layer="tha_admbnda_adm1_rtsd_20220121")
```

We examine the loaded data to confirm the load has been done properly and to get some initial observations of the data.

::: panel-tabset
#### Calling Object

```{r}
thsab_prov
```

#### Checking crs information with st_crs()

```{r}
st_crs(thsab_prov)
```
:::

The output confirms that we have a multipolygon sf object with 77 rows and 17 columns. There is a column named `ADM1_EN` which appears to contain the province names needed to define the BMR boundaries. It also shows that the dataset is using a coordinate reference system rather than a projected reference system.

First, we reload the data to use a projected reference system and apply the correct reference system with EPSG code of 24048 using `st_transform()`. This transformation can be confirmed with `st_crs()` The **tmap** package is then used to visualize the object to see if it properly depicts the boundaries of Thailand and its provinces.

::: panel-tabset
#### Load Object and Transform CRS information

```{r}
thsab_prov <- st_read(dsn="data/geospatial",
                          layer="tha_admbnda_adm1_rtsd_20220121") %>%
  st_transform(crs = 24048)
```

#### Checking crs information with st_crs()

```{r}
st_crs(thsab_prov)
```

#### Plot of thsab_prov using tmap

```{r}
tm_shape(thsab_prov) +
  tm_polygons("grey")
```
:::

## B.2 Filtering THSAB for the Bangkok Metropolitan Region

Before further analyzing the data, we will limit the scope to only consider the Bangkok Metropolitan Region or BMR. This would encompass Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, Samut Sakhon. While it is good to get insights outside of BMR, it is out of the study scope and it is best to focus on the objectives.

The code chunk below checks if all the provinces in the BMR appear as is under the `ADM1_EN` column of `thsab_prov`

```{r}
filter(thsab_prov, ADM1_EN %in% c("Bangkok", "Nonthaburi","Nakhon Pathom", "Pathum Thani", "Samut Prakan", "Samut Sakhon"))$ADM1_EN
```

With the previous code returning all 6 provinces, we have confirmation that the provinces are all present and spelled as is in the data source. We create a new object `bmr_boundary` to contain only the provinces in BMR.

::: panel-tabset
#### Create BMR boundary object using filter()

```{r}
bmr_boundary <- st_read(dsn="data/geospatial",
                      layer="tha_admbnda_adm1_rtsd_20220121") %>%
  st_transform(crs = 24048) %>%
  filter(ADM1_EN %in% c("Bangkok", "Nonthaburi","Nakhon Pathom", "Pathum Thani", "Samut Prakan", "Samut Sakhon"))
  
```

#### Plot of bmr_boundary using tmap

```{r}
tm_shape(bmr_boundary) +
  tm_polygons("grey")
```
:::

The plot confirms that the object. We now need to prepare the .

```{r}
bmr_full = st_union(bmr_boundary)
```

```{r}
tm_shape(bmr_full) +
  tm_polygons("grey")
```

## B.3 THOSM (Thailand Roads Open StreetMap, Shapefile)

The second geospatial object is the street map shapefile. We will use the object `network` to contain the final road network for the study.

```{r}
network <- st_read(dsn="data/geospatial",
                      layer="hotosm_tha_roads_lines_shp")
```

The above code confirms that the dataset is in multilinestring sf format and contains 2.8M records. It also shows that there is no CRS applied to the dataset.

Based on these, the following steps need to be done: apply the right CRS/EPSG code of 24048, and, filter the network to only include BMR. We are opting to filter the network dataset before transforming into a ppp object to limit the amount of resources required as we do not keep the full dataset in memory.

```{r}
network <- st_read(dsn="data/geospatial",
                      layer="hotosm_tha_roads_lines_shp") %>%
  st_set_crs(24048)

```

```{r}
glimpse(network)
```

XXX

## B.4 THRA (Thailand Road Accident Data, Aspatial, csv)

WIP

## B.4 Joining the Aspatial and Geospatial Data

WIP
