---
title: "Discovering Impact of COVID-19 on Thai Tourism Economy"
author: "Federico Jose Rodriguez"
date: "Sep 23 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

(Exercise Summary)

# A. Getting Started

## A.1 Background

Tourism is a major industry in Thailand as it made up to 20% of their gross domestic product pre-pandemic. However, like the rest of the world, the industry has taken a hit with COVID-19 in 2020, and has slowly been recovering since 2021. Recent reports are stating that Thailand is already, but still, at 80% of its peak level in 2019.

While we speak about the industry in general, the state of tourism within Thailand, and their recovery status are not the same. For example, tourism revenues have been focused on Bangkok, Phuket and Chonburi pre-pandemic.

We are interested in understanding the state of tourism across Thailand with regards to its spatial distribution and time and space distribution– both in absolutes and in terms of the trend with respect to the pandemic.

## A.2 Objectives

For this study, we want to understand the state of tourism in Thailand at a provincial level, and answer the following questions:

1.  Are the key tourism indicators in Thailand (at a province level) independent from space and from space and time?

2.  If tourism or any tourism indicators are not independent, what are the clusters, outliers and emerging hotspots and coldspots?

We will use the appropriate packages in R in order to perform the different analysis (spatial and otherwise) to support our answers to the above questions.

## A.3 Data Sources

The following data sources are used for this analysis:

-   [Thailand Domestic Tourism Statistics](https://www.kaggle.com/datasets/thaweewatboy/thailand-domestic-tourism-statistics) from Kaggle covering the years 2019-2023 and are at province and month level across 8 indicators:

    1.  `no_tourist_all` - total number of domestic tourists

    2.  `no_tourist_foreign` - number of foreign tourists

    3.  `no_tourist_occupied` - number of hotel rooms occupied

    4.  `no_tourist_thai` - number of Thai tourists

    5.  `occupancy_rate` - the percentage of occupied travel accommodations (hotel rooms)

    6.  `revenue_all` - total tourism revenue, in M-THB (appears as net profit in the raw data)

    7.  `revenue_foreign` - revenue generated by foreign tourists, in M-THB (appears as net profit in the raw data)

    8.  `revenue_thai` - revenue generated by Thai tourists, in M-THB (appears as net profit in the raw data)

-   [Thailand-Subnational Administrative Boundaries](https://data.humdata.org/dataset/cod-ab-tha?) from Human Data Exchange in shapefile format

## A.4 Importing and Launching R Packages

For this study, the following R packages will be used. A description of the packages and the code, using `p_load()` of the **pacman** package, to import them is given below.

::: panel-tabset
#### Package Description

The loaded packages include:

-   **sf -** package for importing, managing and processing vector-based geospatial data

-   **tidyverse -** collection of packages for performing data importation, wrangling and visualization

-   **tmap -** package with functions for plotting cartographic quality maps

-   **sfdep** - for handling spatial data

-   **coorplot, ggpubr, heatmaply, factoextra** - packages for multivariate data visualization and analysis

-   **cluster, ClustGeo, NbClust** - packages for performing cluster analysis

#### Import Code

```{r}
pacman::p_load(sf, tmap, sfdep, tidyverse,
               ggpubr, heatmaply, factoextra,
               NbClust, cluster, ClustGeo)
```
:::

As we will be performing simulations in the analysis later, it is good practice to define a random seed to be used so that results are consistent for viewers of this report, and the results can be reproduced.

```{r}
set.seed(1234)
```

# B. Data Loading and Preparation

## B.1. Thailand Subnational Boundary, Provincial Level

We load the Thailand subnational administrative boundary shapefile into an R dataframe using `st_read()` from the sf package. We need to analyze at the provincial level so we will be using the files suffixed by "1".

```{r}
thai_sf <- st_read(dsn="data/geospatial", 
                   layer="tha_admbnda_adm1_rtsd_20220121")
```

The output states that the object is of multipolygon geometry type containing 77 features (provinces, records) across 16 fields. (columns) We can check the contents of the object using a number of methods. For the code chunk below, we use `glimpse()` which lists the columns, gives the data type and the first elements.

```{r}
glimpse(thai_sf)
```

For clarity, we can clean up this dataframe by:

1.  Keeping only relevant columns: The province name and code, geometry
2.  Renaming the columns: change ADM1 to Province

The following code chunk executes these steps by using `select()` for the first step and `rename()` for the second step. We again use `glimpse()` to give a preview of the dataset's columns.

```{r}
thai_sf <- thai_sf %>%
  select(ADM1_EN, ADM1_PCODE, geometry) %>%
  rename(Province = ADM1_EN, ProvCode = ADM1_PCODE)

glimpse(thai_sf)
```

We can check if there are any missing values by using `is.na()` and then check across each column using `colSums()` from **Base R**.

```{r}
colSums(is.na(thai_sf))
```

The output shows that there are no missing values for any of the retained columns.

Finally, we can quickly check if the object depicts Thailand properly by producing a quick map using `qtm()` from **tmap** package.

```{r}
qtm(thai_sf)
```

## B.2. Thailand Tourism Data by Province, Jan 2019 - Feb 2023

The code chunk below loads the tourism statistics data into a dataframe `tourism`. We use read_csv() to import the data from the file.

```{r}
tourism <- read_csv("data/aspatial/thailand_domestic_tourism_2019_2023.csv")
```

We can check the contents by using the code chunk below.

```{r}
tourism
```

The imported data contains 7 fields and 30,800 records at a province and month level.

Before we analyze the dataset, let use remove unnecessary columns and rename the column names, similar to the previous dataset, using the code chunk below. (by using `select()` and `rename()`)

```{r}
tourism <- tourism %>%
  select(date, province_eng, region_eng, variable, value) %>%
  rename(Date = date, Province = province_eng, Region = region_eng, Indicator = variable, Value = value)

head(tourism)
```

We have kept only five of the columns which provides the date, the English descriptions for the location (province and region) as well as the (potential) tourism indicator and its value.

We can also check for any missing values across these five columns using the code below. (using `is.na()` and `colSums()`)

```{r}
colSums(is.na(tourism))
```

Each province will be repeated across multiple dates and across multiple indicators. Let us first doublecheck the different values in `Indicator`. We use `unique()` in the code chunk below to achieve this.

```{r}
unique(tourism$Indicator)
```

We are aware that the 'net_profit' indicators are actually revenue so it is better to update them now to avoid misunderstanding later. We use `recode()` from **dplyr** to replace instances with alternative values.

```{r}
tourism <- tourism %>%
  mutate(Indicator = recode(Indicator,
                            "net_profit_all" = "revenue_all",
                            "net_profit_thai" = "revenue_thai",
                            "net_profit_foreign" = "revenue_foreign"))

unique(tourism$Indicator)
```

We will not define which indicators to use until we perform some EDA (Exploratory Data Analysis) in the next section.

Before we move to the next section, we will also introduce some columns into the dataset to make filtering and other analysis easier. For now, we will do this by adding columns for the months and years based on the `Date` column.

```{r}
tourism <- tourism %>%
  mutate(Year = year(Date)) %>%
  mutate(MonthNum = month(Date)) %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  mutate(MonthYear = format(ymd(Date), "%Y-%m"))

head(tourism)
```

# C. Exploratory Data Analysis

We will use data analysis and visualization techniques to investigate the data before we perform the main analyses required. This allows us to get some insights on things like variables to use, provinces to focus on, deep dives to perform etc.

## C.1 Tourism Revenue

We first look at tourism revenue which is currently reported in million Thai baht. We will use a constant rate of 34.784 THB per USD based on 2023 exchange rates to scale down the numbers and transform it into something more recognizable for most of the readers.

We first create a plot for the monthly tourism revenue in total and by foreign and local tourists. The code below selects the relevant data and prepares the line plot using `ggplot()`. Finally, we use `ggplotly()` to render it as an interactive chart so we can easily examine the resulting chart.

```{r}
# Subset the data to just the required indicators
aggregated_data <- tourism %>%
  filter(Indicator %in% c("revenue_all", "revenue_thai", "revenue_foreign")) %>%
  group_by(MonthYear, Indicator) %>%
  summarise(TotalValue = sum(Value, na.rm = TRUE) / 34.784) %>%
  ungroup()

# Create the line chart
p <- ggplot(aggregated_data, aes(x = MonthYear, y = TotalValue, color = Indicator, group = Indicator)) +
  geom_line(size = 1) +
  labs(title = "Thai Tourism Revenue by Month",
       y = "Million USD",
       x = "Month-Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("revenue_all" = "blue", "revenue_thai" = "green", "revenue_foreign" = "red"))

# Convert the ggplot chart to an interactive plotly chart
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot
```

The resulting chart is consistent with the expectation on the impact and recovery from the pandemic. We can view the above chart as a timeline:

1.  Up to (mid)January 2020: Pre-covid. No travel restrictions have been set yet. (Jan 2019 - Jan 2020, 13 months)

2.  Feb 2020 to November 2021: Covid. Various lockdown measures in place. All foreign non-essential travel is banned. There is some local tourist activity, but another set of measures in May 2021 again prevents non-essential movement (Feb 2020 - Oct 2021, 22 months)

3.  November 2021 onwards: Post-Covid. Travel restrictions have been eased or lifted and tourism revenues have been recovering (Nov 2021 - Feb 2023, 16 months)

Pre- and post-covid, we see that foreign tourists contribute more to the overall revenue, and their contribution has a large amount of variance. Local tourists during the same period have contributed a more stable amount month-on-month.

We can code the three periods mentioned above into the tourism dataset for convenience. We use the `ifelse()` function to do this based on the cutoff dates mentioned above.

```{r}
tourism$Period <- ifelse(tourism$Date < as.Date("2020-02-01"), "Pre-Covid",
                         ifelse(tourism$Date > as.Date("2021-10-01"), "Post-Covid", "Covid"))
```

We can next check the tourism revenue at the province level. As plotting all 77 provinces across all the periods will not produce readable charts, we will focus on top 20 provinces for the different periods. We will also take the average monthly revenue rather than the total since each period has a different number of months.

We prepare a new dataframe that summarizes the indicators for each province in `tourism`. Aside from the average revenue per period, we will also compute for the average number of visitors as well as the average spend per visitor.

```{r}
tourism_period <- tourism %>%
  group_by(Province) %>%
  summarise(
    PreCovid_Revenue_total = sum(Value[Period == "Pre-Covid" & Indicator == "revenue_all"], na.rm = TRUE)/13/34.784,
    Covid_Revenue_total = sum(Value[Period == "Covid" & Indicator == "revenue_all"], na.rm = TRUE)/22/34.784,
    PostCovid_Revenue_total = sum(Value[Period == "Post-Covid" & Indicator == "revenue_all"], na.rm = TRUE)/16/34.784,
    PreCovid_Revenue_foreign = sum(Value[Period == "Pre-Covid" & Indicator == "revenue_foreign"], na.rm = TRUE)/13/34.784,
    Covid_Revenue_foreign = sum(Value[Period == "Covid" & Indicator == "revenue_foreign"], na.rm = TRUE)/22/34.784,
    PostCovid_Revenue_foreign = sum(Value[Period == "Post-Covid" & Indicator == "revenue_foreign"], na.rm = TRUE)/16/34.784,
    PreCovid_Revenue_thai = sum(Value[Period == "Pre-Covid" & Indicator == "revenue_thai"], na.rm = TRUE)/13/34.784,
    Covid_Revenue_thai = sum(Value[Period == "Covid" & Indicator == "revenue_thai"], na.rm = TRUE)/22/34.784,
    PostCovid_Revenue_thai = sum(Value[Period == "Post-Covid" & Indicator == "revenue_thai"], na.rm = TRUE)/16/34.784,
    PreCovid_tourists_total = sum(Value[Period == "Pre-Covid" & Indicator == "no_tourist_all"], na.rm = TRUE)/13,
    Covid_tourists_total = sum(Value[Period == "Covid" & Indicator == "no_tourist_all"], na.rm = TRUE)/22,
    PostCovid_tourists_total = sum(Value[Period == "Post-Covid" & Indicator == "no_tourist_all"], na.rm = TRUE)/16,
    PreCovid_tourists_foreign = sum(Value[Period == "Pre-Covid" & Indicator == "no_tourist_foreign"], na.rm = TRUE)/13,
    Covid_tourists_foreign = sum(Value[Period == "Covid" & Indicator == "no_tourist_foreign"], na.rm = TRUE)/22,
    PostCovid_tourists_foreign = sum(Value[Period == "Post-Covid" & Indicator == "no_tourist_foreign"], na.rm = TRUE)/16,
    PreCovid_tourists_thai = sum(Value[Period == "Pre-Covid" & Indicator == "no_tourist_thai"], na.rm = TRUE)/13,
    Covid_tourists_thai = sum(Value[Period == "Covid" & Indicator == "no_tourist_thai"], na.rm = TRUE)/22,
    PostCovid_tourists_thai = sum(Value[Period == "Post-Covid" & Indicator == "no_tourist_thai"], na.rm = TRUE)/16
  ) %>%
  mutate(PreCovidSpend_total = PreCovid_Revenue_total / PreCovid_tourists_total * 1000000) %>%
  mutate(CovidSpend_total = Covid_Revenue_total / Covid_tourists_total * 1000000) %>%
  mutate(PostCovidSpend_total = PostCovid_Revenue_total / PostCovid_tourists_total * 1000000) %>%
  mutate(PreCovidSpend_foreign = PreCovid_Revenue_foreign / PreCovid_tourists_foreign * 1000000) %>%
  mutate(CovidSpend_foreign = Covid_Revenue_foreign / Covid_tourists_foreign * 1000000) %>%
  mutate(PostCovidSpend_foreign = PostCovid_Revenue_foreign / PostCovid_tourists_foreign * 1000000) %>%
  mutate(PreCovidSpend_thai = PreCovid_Revenue_thai / PreCovid_tourists_thai * 1000000) %>%
  mutate(CovidSpend_thai = Covid_Revenue_thai / Covid_tourists_thai * 1000000) %>%
  mutate(PostCovidSpend_thai = PostCovid_Revenue_thai / PostCovid_tourists_thai * 1000000)
  
```

With the summarized dataframe prepared, we can now prepare a few visualizations to look at the provinces with regards to the average monthly tourism revenue.

First, let us try using a scatterplot to see both the average revenue pre-Covid (x-axis) and post-Covid. (y-axis) Provinces with the highest pre-Covid revenue will appear the rightmost, while those that have the highest post-Covid revenue will appear the rightmost.

The code below uses the **plotly** package to produce an interactive scatterplot of the pre- and post-covid average monthly revenue for all tourists. With the interactive chart, the province names will be visible by hovering over and the user can zoom in by creating a selection in the chart.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovid_Revenue_total,
  y = ~PostCovid_Revenue_total,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Revenue - M-USD',
    xaxis = list(title = 'Average PreCovid Revenue'),
    yaxis = list(title = 'Average PostCovid Revenue')
  )

# Display the plot
plot
```

Baed on the plot, we see that Bangkok, Phuket and Chonburi have consistently been the top 3 highest revenue generating before and after the pandemic. When we look further down the list, we see some shifts for some of the provinces.

To aid the reader, we recreate the chart with those top 3 provinces excluded using the code chunk below.

```{r}
plot <- plot_ly(
  data = filter(tourism_period, !(Province %in% c("Bangkok", "Phuket", "Chonburi"))),
  x = ~PreCovid_Revenue_total,
  y = ~PostCovid_Revenue_total,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Revenue - M-USD, exc Top 3 Provinces',
    xaxis = list(title = 'Average PreCovid Revenue'),
    yaxis = list(title = 'Average PostCovid Revenue')
  )

# Display the plot
plot
```

Some key observations from the above charts are:

1.  Chiang Mai has moved from top 5 to top 4. A large reason for this is a drop from Krabi. Pre-covid, Krabi was top 4, but has dropped to at least top 10.

2.  Chiang Rai and Prachuap Khiri Khan have risen to top 5 and 6. These provinces were top 9 or lower before.

3.  Songkhla and Phang Nga were in the top 10 pre-Covid but are also showing a drop in ranking post-Covid

We can do the same chart for just the revenue from foreign tourists using the code chunk below.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovid_Revenue_foreign,
  y = ~PostCovid_Revenue_foreign,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Revenue from Foreign Tourists - M-USD',
    xaxis = list(title = 'Average PreCovid Revenue'),
    yaxis = list(title = 'Average PostCovid Revenue')
  )

# Display the plot
plot
```

The top 3 provinces are the same for both, but there are differences further down the list.

We can also show these numbers graphically in a map. Before we do this, let us add one set of measures in tourism_period to indicate the recovery rate. This will be the ratio of the post-covid and pre-covid measures and will indicate how much of the pre-covid level has been achieved. (on average)

We perform this using the code chunk below. We do this for all sets of measures across revenue, number of tourists and the average spending.

```{r}
tourism_period <- tourism_period %>%
  mutate(Revenue_total_recovery = PostCovid_Revenue_total / PreCovid_Revenue_total) %>%
  mutate(Revenue_foreign_recovery = PostCovid_Revenue_foreign / PreCovid_Revenue_foreign) %>%
  mutate(Revenue_thai_recovery = PostCovid_Revenue_thai / PreCovid_Revenue_thai) %>%
  mutate(Tourists_total_recovery = PostCovid_tourists_total / PreCovid_tourists_total) %>%
  mutate(Tourists_foreign_recovery = PostCovid_tourists_foreign / PreCovid_tourists_foreign) %>%
  mutate(Tourists_thai_recovery = PostCovid_tourists_thai / PreCovid_tourists_thai) %>%
  mutate(Spend_total_recovery = PostCovidSpend_total / PreCovidSpend_total) %>%
  mutate(Spend_foreign_recovery = PostCovidSpend_foreign / PreCovidSpend_foreign) %>%
  mutate(Spend_thai_recovery = PostCovidSpend_thai / PreCovidSpend_thai)
```

We need to include all the indicators into the sf dataframe. This means merging the `tourism_period` and the `thai_sf` dataframes. Let us first check that the naming is the same for both dataframes by checking which values do not have a match. We use the code below which uses `left_join()` to match and then `filter()` to check those that do not have matches.

```{r}
# Identify mismatched Province names in tourism_period
mismatched_values <- tourism_period %>%
  left_join(thai_sf, by = "Province") %>%
  filter(is.na(ProvCode)) %>%
  select(Province)

mismatched_tourism <- mismatched_values$Province

# Identify mismatched Province names in thai_sf
mismatched_values <- thai_sf %>%
  left_join(tourism_period, by = "Province") %>%
  filter(is.na(Covid_Revenue_total)) %>%
  select(Province)

mismatched_thai <- mismatched_values$Province

# Print the mismatched values
list(
  mismatched_in_tourism_period = mismatched_tourism,
  mismatched_in_thai_sf = mismatched_thai
)

```

We see that there are 8 mismatched province names for each of the dataframes. We need to standardize these namings to ensure that the indicators are mapped to the correct province. We will opt to keep the descriptions from tourism_period which gives more compact naming. We use `recode()` in the code chunk below to accomplish this in a new dataframe.

```{r}
thaitourism_sf <- thai_sf %>%
  mutate(Province = recode(Province,
                            "Lop Buri" = "Lopburi",
                            "Chai Nat" = "Chainat",
                            "Chon Buri" = "Chonburi",
                            "Prachin Buri" = "Prachinburi",
                            "Buri Ram" = "Buriram",
                            "Si Sa Ket" = "Sisaket",
                            "Nong Bua Lam Phu" = "Nong Bua Lamphu",
                            "Phangnga" = "Phang Nga"))
```

We can now use `leftjoin()` in the codechunk below to merge the two datasets.

```{r}
thaitourism_sf <- left_join(thaitourism_sf, tourism_period,
                     by=c("Province"="Province"))
```

The code chunk below confirms that the new object is still an sf dataframe.

```{r}
class(thaitourism_sf)
```

We can use the tmap package to produce side-by side maps of Thailand with the average monthly tourism revenue before and after covid using the code chunk below. Given the wide range of values, we will use quantiles for the data classes. We also include the recovery rate of each of the provinces as a third map.

```{r, fig.width=15, fig.height=8}
tm_shape(thaitourism_sf) +
  tm_fill(col = c("PreCovid_Revenue_total", "PostCovid_Revenue_total", "Revenue_total_recovery"),
          style = "quantile",
          palette = list("viridis", "viridis", "RdYlGn"),
          title = c("Monthly Revenue - M-USD", "Monthly Revenue - M-USD", "Post vs Pre")) +
  tm_borders(col = "black") +
  tm_layout(title = c("Pre Covid", "Post Covid", "Recovery Rate"),
            title.position = c("right", "top"),
            legend.position = c("right", "bottom"),
            bg.color = "grey90")

```

As also seen in the scatterplot, we also see some change in rankings with the maps. For example, some provinces previously in the top 20% have moved down to the next 20%. (e.g., Khon Kaen and Phang Nga)

If we focus on the third map, we also see what seems like a cluster of provinces in the south which are lagging with regards to their recovery on the average tourism revenue. We will watch out for these once we conduct our cluster analyses.

## C.2 Number of Tourists

The next measure we can look at is the number of tourists. We produce a similar line chart as we did for tourism revenue with the code chunk below. We display the number of tourists in thousands.

```{r}
# Subset the data to just the required indicators
aggregated_data <- tourism %>%
  filter(Indicator %in% c("no_tourist_all", "no_tourist_thai", "no_tourist_foreign")) %>%
  group_by(MonthYear, Indicator) %>%
  summarise(TotalValue = sum(Value, na.rm = TRUE) / 1000) %>%
  ungroup()

# Create the line chart
p <- ggplot(aggregated_data, aes(x = MonthYear, y = TotalValue, color = Indicator, group = Indicator)) +
  geom_line(size = 1) +
  labs(title = "Thailand Number of Tourists by Month",
       y = "Tourists, Thousands",
       x = "Month-Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("no_tourist_all" = "blue", "no_tourist_thai" = "green", "no_tourist_foreign" = "red"))

# Convert the ggplot chart to an interactive plotly chart
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot
```

The trend for the total follows the same general movement as the chart for revenue, however, it looks like tourist numbers are primarily driven by locals than foreigners.

Similar to the previous section, we can produce an interactive scatterplot to see the number of tourists each province gets on average before and after the pandemic. We do this first for the total number of tourists.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovid_tourists_total,
  y = ~PostCovid_tourists_total,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid No of Tourists',
    xaxis = list(title = 'PreCovid Average'),
    yaxis = list(title = 'PostCovid Average')
  )

# Display the plot
plot
```

Bangkok, Chonburi and Phuket had the highest number of visitors pre-Covid, but Phuket appears to have dropped off in favor of Kanchanaburi post-covid.

We can also look at the foreign tourists using the code chunk below. We skip local tourists and focus only on foreign ones as they deviate from the overall number.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovid_tourists_foreign,
  y = ~PostCovid_tourists_foreign,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid No of Foreign Tourists',
    xaxis = list(title = 'PreCovid Average'),
    yaxis = list(title = 'PostCovid Average')
  )

# Display the plot
plot
```

Bangkok, Phuket and Chonburi appear as the top 3 destinations for foreign tourists (in terms of number) before and after Covid. We also see that Krabi dropped from the top 3 post covid.

We can also produce side-by-side maps for the number of tourists and their recovery rates using the code chunk below.

```{r, fig.width=15, fig.height=8}
tm_shape(thaitourism_sf) +
  tm_fill(col = c("PreCovid_tourists_total", "PostCovid_tourists_total", "Tourists_total_recovery"),
          style = "quantile",
          palette = list("viridis", "viridis", "RdYlGn"),
          title = c("Monthly Tourists", "Monthly Tourists", "Post vs Pre")) +
  tm_borders(col = "black") +
  tm_layout(title = c("Pre Covid", "Post Covid", "Recovery Rate"),
            title.position = c("right", "top"),
            legend.position = c("right", "bottom"),
            bg.color = "grey90")

```

We again see the southern region mostly lagging with regards to their recovery post covid also in terms of the number of visitors.

## C.3 Average Spend Per Visitor

We next look at the average per spend per visitor which is the quotient of the tourism revenue and the total number of tourists. This will tell us whether tourists are spending more or less around the pandemic.

First we produce a similar line graph as before to look at the trend at an overall picture using the code chunk below.

```{r}
# Subset the data to just the required indicators
aggregated_data <<- tourism %>%
  group_by(MonthYear) %>%
  summarise(
    Spend_total = sum(Value[Indicator == "revenue_all"]) / sum(Value[Indicator == "no_tourist_all"]) * 1000000 / 34.784,
    Spend_thai = sum(Value[Indicator == "revenue_thai"]) / sum(Value[Indicator == "no_tourist_thai"]) * 1000000 / 34.784,
    Spend_foreign = sum(Value[Indicator == "revenue_foreign"]) / sum(Value[Indicator == "no_tourist_foreign"]) * 1000000 / 34.784
  ) %>%
  pivot_longer(cols = starts_with("Spend"), names_to = "Indicator", values_to = "TotalValue") %>%
  mutate(Indicator = case_when(
    Indicator == "Spend_total" ~ "Spend_total",
    Indicator == "Spend_thai" ~ "Spend_thai",
    Indicator == "Spend_foreign" ~ "Spend_foreign"
  ))

# Create the line chart
p <- ggplot(aggregated_data, aes(x = MonthYear, y = TotalValue, color = Indicator, group = Indicator)) +
  geom_line(size = 1) +
  labs(title = "Thailand Average Spend Per Tourist by Month",
       y = "Average Spend, USD",
       x = "Month-Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom") +
  scale_color_manual(values = c("Spend_total" = "blue", "Spend_thai" = "green", "Spend_foreign" = "red"))

# Convert the ggplot chart to an interactive plotly chart
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot
```

The chart shows that the average spend of foreign tourists are much higher than local ones and appears to be the same before and after Covid. There appears to be a shift in the average spend for all tourists overall which is probably driven by an increase in the contribution for the number of local tourists versus foreign tourists. Local tourists appear to show a step decrease after covid as well in terms of their average spending.

Next, we check the average spending of tourists across provinces using a similar scatterplot as before.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovidSpend_total,
  y = ~PostCovidSpend_total,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Average Spend Per Tourist',
    xaxis = list(title = 'PreCovid Average $'),
    yaxis = list(title = 'PostCovid Average $')
  )

# Display the plot
plot
```

Phuket saw the highest average spend pre- and post-covid. Krabi was second highest pre-Covid but drop to third, as Bangkok rose from fourth to second over that period.

We can do the same for foreign tourists as the earlier chart showed that it was very different from the total. We use the code chunk below tor produce a similar scatterplot but taking the foreign tourist figures.

```{r}
plot <- plot_ly(
  data = tourism_period,
  x = ~PreCovidSpend_foreign,
  y = ~PostCovidSpend_foreign,
  type = 'scatter',
  mode = 'markers',
  text = ~Province,
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Average Spend Per Foreign Tourist',
    xaxis = list(title = 'PreCovid Average $'),
    yaxis = list(title = 'PostCovid Average $')
  )

# Display the plot
plot
```

Phuket, Bangkok and Chonburi have been consistent as top 4 highest spend for foreign tourists. From the chart above, we see that Chiang Rai has risen to the number five spot for highest foreign tourist spending post-Covid.

We can also produce a map visualization of the average spend using **tmap** package. We focus on the total tourist population in the visualization below.

```{r, fig.width=15, fig.height=8}
tm_shape(thaitourism_sf) +
  tm_fill(col = c("PreCovidSpend_total", "PostCovidSpend_total", "Spend_total_recovery"),
          style = "quantile",
          palette = list("viridis", "viridis", "RdYlGn"),
          title = c("Average Tourist Spending", "Average Tourist Spending", "Post vs Pre")) +
  tm_borders(col = "black") +
  tm_layout(title = c("Pre Covid", "Post Covid", "Recovery Rate"),
            title.position = c("right", "top"),
            legend.position = c("right", "bottom"),
            bg.color = "grey90")
```

We again see slow recovery in the southern region, but at the same time, this region has the highest average spending pre- and post-covid. (i.e., top 20%)

## C.4 Occupancy Rate

The final single indicator we will look at is the occupancy rate. We have not prepared and included the data for occupancy rate before as this is a ratio measure which be just summed or averaged. In order to be able to aggregate occupancy rate, we not only need the actual occupancy rate from the data, but we also need the number of rooms occupied which is given by the `no_tourist_occupied` indicator in the data, and also the number of rooms in total– which is not included in the data.

The following code chunk prepares a new dataframe from tourist with the following transformation steps:

1.  Retain MonthYear, Province, and records for occupany rate and number of rooms occupied

2.  Keep values for the two indicators as separate columns. We use average just in case a province appears multiple times on the same date (to resolve conflicts)

3.  We compute the total number of rooms as the number of rooms occupied divided by the occupancy rate

4.  We add the tags for the period for pre- and post- covid

```{r}
occupancy_df <- tourism %>%
  filter(Indicator %in% c("occupancy_rate", "no_tourist_occupied")) %>%
  group_by(Date, MonthYear, Province) %>%
  summarise(
    occupancy = mean(Value[Indicator == "occupancy_rate"], na.rm = TRUE),
    occupied_rooms = mean(Value[Indicator == "no_tourist_occupied"], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(total_rooms = ifelse(occupancy == 0, 0, occupied_rooms / occupancy * 100))

occupancy_df$Period <- ifelse(occupancy_df$Date < as.Date("2020-02-01"), "Pre-Covid",
                         ifelse(occupancy_df$Date > as.Date("2021-10-01"), "Post-Covid", "Covid"))

head(occupancy_df)
```

Before we produce the charts, let us update `thaitourism_sf` with the aggregated occupancy rates by:

1.  Computing Pre- and Post-covid total number of rooms and occupied rooms per province
2.  Computing Pre- and Post-covid occupancy rate per province based on step 1
3.  Add the new columns into `thaitourism_sf` using `left_join()`

The first two steps are accomplished in the first code block while the third step is accomplished in the second.

```{r}
occupancy_summary <- occupancy_df %>%
  group_by(Province) %>%
  summarise(
    PreCovid_occupied_rooms = sum(occupied_rooms[Period == "Pre-Covid"], na.rm = TRUE),
    PostCovid_occupied_rooms = sum(occupied_rooms[Period == "Post-Covid"], na.rm = TRUE),
    PreCovid_total_rooms = sum(total_rooms[Period == "Pre-Covid"], na.rm = TRUE),
    PostCovid_total_rooms = sum(total_rooms[Period == "Post-Covid"], na.rm = TRUE)
  ) %>%
  mutate(
    PreCovid_occupancy = ifelse(PreCovid_total_rooms == 0, 0, (PreCovid_occupied_rooms / PreCovid_total_rooms) * 100),
    PostCovid_occupancy = ifelse(PostCovid_total_rooms == 0, 0, (PostCovid_occupied_rooms / PostCovid_total_rooms) * 100)
  )

head(occupancy_summary)
```

```{r}
thaitourism_sf <- left_join(thaitourism_sf, occupancy_summary, by = c("Province"="Province"))

class(thaitourism_sf)
```

Finally, we add a column for the recovery of the occupancy rate using the code chunk below.

```{r}
thaitourism_sf <- mutate(thaitourism_sf, Occupancy_recovery = ifelse(PreCovid_occupancy == 0, 0, (PostCovid_occupancy / PreCovid_occupancy)), .before = -1)
```

```{r}
head(thaitourism_sf)
```

For the first visualization, let us plot the occupancy rate at a total level. The code below summarizes based on `occupancy_df` and computes a national occupancy rate to plot in a line chart.

```{r}
aggregated_data <<- occupancy_df %>%
  group_by(MonthYear) %>%
  summarise(
    occupancy_rate = sum(occupied_rooms, na.rm = TRUE) / sum(total_rooms, na.rm = TRUE) * 100
  )

ggplot(aggregated_data, aes(x = MonthYear, y = occupancy_rate, group = 1)) +
  geom_line() +
  labs(
    title = "Occupancy Rate Over Time",
    x = "MonthYear",
    y = "Occupancy Rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

The chart shows that occupancy rate has picked up after October 2021 and appears to have more or less reached pre-covid levels in the most recent months.

We can also produce a scatterplot to show pre- and post-covid occupancy rates at a province level. We do this using the code chunk below which produces an interactive plot.

```{r}
plot <- plot_ly(
  data = occupancy_summary,
  x = ~PreCovid_occupancy,
  y = ~PostCovid_occupancy,
  type = 'scatter',
  mode = 'markers',
  text = ~paste('Province:', Province, '<br>PreCovid Occupancy:', round(PreCovid_occupancy), '<br>PostCovid Occupancy:', round(PostCovid_occupancy)),
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Monthly Pre- and Post-Covid Occupancy Rate (%)',
    xaxis = list(title = 'PreCovid Occupancy'),
    yaxis = list(title = 'PostCovid Occupancy')
  )

# Display the plot
plot
```

The occupancy rates are showing a dispersed pattern as the overall rankings on occupancy rates have significantly changed pre- and post-covid. Bangkok, Chonburi and Suphan Buri reported the highest occupancy rates before covid, but Nan, Chang Rai and Nakhon Phanom have the highest rates post covid.

The next visualization for this measure is a side-by-side map for the pre- and post-covid occupancy rates as well as the recovery rate for occupancy. We use the code chunk below which uses **tmap** package to produce the maps.

```{r, fig.width=15, fig.height=8}
tm_shape(thaitourism_sf) +
  tm_fill(col = c("PreCovid_occupancy", "PostCovid_occupancy", "Occupancy_recovery"),
          style = "quantile",
          palette = list("viridis", "viridis", "RdYlGn"),
          title = c("Average Occupancy Rate", "Average Occupancy Rate", "Post vs Pre")) +
  tm_borders(col = "black") +
  tm_layout(title = c("Pre Covid", "Post Covid", "Recovery Rate"),
            title.position = c("right", "top"),
            legend.position = c("right", "bottom"),
            bg.color = "grey90")
```

Before we leave this section, let us try to understand occupancy rate a bit more. Going back to earlier, we want to understand if high occupancy rate post-Covid is being driven by the number of available rooms. To help us answer this, we create a scatterplot of the available rooms against the occupancy rate post-Covid using the code chunk below.

```{r}
plot <- plot_ly(
  data = occupancy_summary,
  x = ~PostCovid_total_rooms,
  y = ~PostCovid_occupancy,
  type = 'scatter',
  mode = 'markers',
  text = ~paste('Province:', Province, '<br>Number of Rooms:', round(PostCovid_total_rooms), '<br>Occupancy Rate:', round(PostCovid_occupancy)),
  hoverinfo = 'text'
) %>%
  layout(
    title = 'Post-Covid Number of Rooms and Occupancy Rate (%)',
    xaxis = list(title = 'Number of Rooms'),
    yaxis = list(title = 'Occupancy Rate')
  )

# Display the plot
plot
```

Below 10M rooms, there appears to be an upward trend in occupancy rate to the number of rooms. Provinces with more than 10M rooms go against the trend and appear to be capped to 60% occupancy.

# D. Spatial Autocorrelation

Testing for spatial autocorrelation allows us to statistically determine if any of the indicator variables are dependent on space or location. These techniques also are useful for classifying clusters and outliers for particular variables of concern.

## D.1 Variable Selection

We will turn our attention to the first three measures discussed discussed in the previous section: **tourism revenue, number of tourists** and **average tourist spending**. We will not analyse the occupancy rate further as it is highly dependent on the number of rooms.

We will focus on checking signs of spatial autocorrelation or association before and after covid, as well as the recovery rate at the overall level for these three indicators– so we will be looking at 9 variables for our analysis.

## D.2 Deriving the Contiguity and Weight Matrix

For the tests for this section, we need to derive a neighbor list as well as a weight matrix for each province to its neighbors. We do this in the code chunk below which creates a copy of `thai_tourism_sf` and includes a neighbour list `nb` and the weights `wt` using the Queen's criterion and using equal weights. We use the `include_self()` on `st_contiguity()` so that each node will have itself, at the minimum, as its neighbor. We need to ensure that each node has a non-empty neighbor list and this approach still limits neighbors to provinces that share a border.

```{r}
wm_q <- thaitourism_sf %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_weights(nb,
                         style="W"),
         .before=1)
```

## D.3 Global Moran's I Test

Global tests of spatial autocorrelation compares the value of each point/province to the overall value in order to conclude on spatial dependence. For this, we will focus on using Global Moran's I which will work on the following general hypotheses:

-   $H_0$ - The value of (variable) is randomly distributed across provinces in Thailand

-   $H_1$ - The value of (variable) is not randomly distributed across provinces in Thailand

Further, the value of the test statistic $I$ will also give indication on the underlying pattern:

-   $I > 0$ - Clustering; observations tend to be similar

-   $I < 0$ - Dispersed / regular; observations tend to be dissimilar

-   Where $I$ close to zero - observations are arranged randomly

To perform Global Moran's I test, with permutations, we use `global_moran_perm()` from **sfdep** package. We will use a 5% significance level for all the testing to be performed, and we will run 100 permutations / simulations for each test.

### D.3.1 Global Moran's Test on Tourism Revenue

The code chunks in the tabs below run the Global Moran's I permutation test on total tourism revenue (per month) pre-Covid, post-Covid and the recovery rate.

::: panel-tabset
#### Pre-Covid Revenue

```{r}
global_moran_perm(wm_q$PreCovid_Revenue_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Post-Covid Revenue

```{r}
global_moran_perm(wm_q$PostCovid_Revenue_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Revenue Recovery Rate

```{r}
global_moran_perm(wm_q$Revenue_total_recovery,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```
:::

The results show no evidence to reject spatial independence for the total revenue pre- and post-Covid. However, it shows **signs of clustering for the revenue recovery rate** as the p-value is below 0.05 and the statistic is above 1.

### D.3.2 Global Moran's Test on Number of Tourists

The code chunks in the tabs below run the Global Moran's I permutation test on total number of tourists (per month) pre-Covid, post-Covid and the recovery rate.

::: panel-tabset
#### Pre-Covid Number of Tourists

```{r}
global_moran_perm(wm_q$PreCovid_tourists_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Post-Covid Number of Tourists

```{r}
global_moran_perm(wm_q$PostCovid_tourists_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Number of Tourists Recovery Rate

```{r}
global_moran_perm(wm_q$Tourists_total_recovery,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```
:::

We see similar results here. The results show no evidence to reject spatial independence for the total number of tourists pre- and post-Covid. However, it shows **signs of clustering for the number of tourists recovery rate** as the p-value is below 0.05 and the statistic is above 1.

### D.3.3 Global Moran's Test on Average Tourist Spend

The code chunks in the tabs below run the Global Moran's I permutation test on average tourist spend pre-Covid, post-Covid and the recovery rate.

::: panel-tabset
#### Pre-Covid Average Spend

```{r}
global_moran_perm(wm_q$PreCovidSpend_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Post-Covid Average Spend

```{r}
global_moran_perm(wm_q$PostCovidSpend_total,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

#### Average Spend Recovery Rate

```{r}
global_moran_perm(wm_q$Spend_total_recovery,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```
:::

Average tourist spending is showing signs of clustering on all dimensions (pre-Covid, post-Covid and the recovery rate) as the p-value is below 0.05 and the I statistic is anove 0 in all cases.
